---
title: "PLFS Basic Analysis"
subtitle: "Labour Force Indicators from Periodic Labour Force Survey"
author: "Your Name"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
  docx:
    toc: true
    reference-doc: null
  pdf:
    toc: true
    documentclass: article
    geometry: margin=1in
execute:
  echo: true
  warning: false
  message: false
params:
  data_file: "plfs_2022_23_person"  # Parquet file name (without extension)
  survey_year: "2022-23"
  approach: "ps"  # "ps" for Principal Status, "cws" for Current Weekly Status
---

```{r setup}
#| label: setup
#| include: false

# Load all project functions
source(here::here("R/01_config.R"))
source(here::here("R/02_read_microdata.R"))
source(here::here("R/03_survey_design.R"))
source(here::here("R/03_survey_design.R"))
source(here::here("R/04_plfs_indicators.R"))
source(here::here("R/05_codebook_utils.R"))
source(here::here("R/06_export_tables.R"))
source(here::here("R/07_viz_themes.R"))

# Additional packages for this report
library(knitr)
library(kableExtra)

# Set default table format based on output
if (knitr::is_html_output()) {
  options(knitr.table.format = "html")
} else if (knitr::is_latex_output()) {
  options(knitr.table.format = "latex")
} else {
  options(knitr.table.format = "simple")
}
```

# Introduction

This report presents key labour force indicators from the **Periodic Labour Force Survey (PLFS) `r params$survey_year`** conducted by the National Statistical Office (NSO), Ministry of Statistics and Programme Implementation, Government of India.

**Key Indicators Covered:**

- Labour Force Participation Rate (LFPR)
- Worker Population Ratio (WPR)
- Unemployment Rate (UR)
- Employment distribution by type

**Methodology:**

- Activity status approach: `r if(params$approach == "ps") "Principal Status (PS)" else "Current Weekly Status (CWS)"`
- Age group: 15 years and above
- Survey weights applied for population-level estimates
- 95% confidence intervals reported

---

# Data Loading

```{r load-data}
#| label: load-data

# Load processed PLFS data
persons <- load_from_parquet(params$data_file)

# Quick overview
cat(sprintf("Dataset: %s\n", params$data_file))
cat(sprintf("Observations: %s\n", format(nrow(persons), big.mark = ",")))
cat(sprintf("Variables: %d\n", ncol(persons)))
```

```{r decode-data}
#| label: decode-data

# Decode categorical variables
persons <- decode_all(
  persons,
  state_col = "State",
  sex_col = "Sex",
  sector_col = "Sector"
)

# Preview
head(persons[, .(State, State_name, Sex, Sector, Age)], 10) |>
  kable(caption = "Sample of decoded data")
```

---

# Survey Design

```{r create-design}
#| label: create-design

# Create survey design with proper weights and stratification
design <- create_plfs_design(persons, level = "person")

# Summary
survey_design_summary(design)
```

---

# National Indicators

## Overall Labour Force Indicators

```{r national-overall}
#| label: national-overall

# Calculate national indicators
national <- calc_all_indicators(design, approach = params$approach)

# Display
national_display <- data.table(
  Indicator = c("Labour Force Participation Rate", 
                "Worker Population Ratio", 
                "Unemployment Rate"),
  Estimate = c(
    sprintf("%.1f%%", national$lfpr),
    sprintf("%.1f%%", national$wpr),
    sprintf("%.1f%%", national$ur)
  ),
  `Std. Error` = c(
    sprintf("%.2f", national$lfpr_se),
    sprintf("%.2f", national$wpr_se),
    sprintf("%.2f", national$ur_se)
  ),
  `Sample Size` = format(national$n, big.mark = ",")
)

kable(national_display, 
      caption = sprintf("National Labour Force Indicators, PLFS %s", params$survey_year),
      align = c("l", "r", "r", "r"))
```

## By Sex

```{r national-by-sex}
#| label: national-by-sex

by_sex <- calc_indicators_by_sex(design, approach = params$approach)

# Format for display
by_sex_display <- by_sex[, .(
  Sex,
  `LFPR (%)` = sprintf("%.1f (%.1f-%.1f)", lfpr, lfpr_low, lfpr_upp),
  `WPR (%)` = sprintf("%.1f (%.1f-%.1f)", wpr, wpr_low, wpr_upp),
  `UR (%)` = sprintf("%.1f (%.1f-%.1f)", ur, ur_low, ur_upp),
  N = format(n, big.mark = ",")
)]

kable(by_sex_display,
      caption = "Labour Force Indicators by Sex (95% CI)",
      align = c("l", "r", "r", "r", "r"))
```

```{r plot-by-sex}
#| label: plot-by-sex
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Labour Force Participation Rate by Sex"

# Reshape for plotting
by_sex_long <- melt(by_sex, 
                    id.vars = "Sex",
                    measure.vars = c("lfpr", "wpr", "ur"),
                    variable.name = "indicator",
                    value.name = "value")

by_sex_long[, indicator := factor(indicator, 
                                   levels = c("lfpr", "wpr", "ur"),
                                   labels = c("LFPR", "WPR", "Unemployment Rate"))]

ggplot(by_sex_long, aes(x = indicator, y = value, fill = Sex)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = sprintf("%.1f", value)),
            position = position_dodge(width = 0.8),
            vjust = -0.5, size = 3) +
  scale_fill_pub("binary") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = sprintf("Labour Force Indicators by Sex, PLFS %s", params$survey_year),
    subtitle = sprintf("%s approach, Age 15+", 
                       if(params$approach == "ps") "Principal Status" else "CWS"),
    x = NULL,
    y = "Percentage (%)",
    fill = NULL
  ) +
  theme_publication()
```

## By Sector (Rural/Urban)

```{r national-by-sector}
#| label: national-by-sector

by_sector <- calc_all_indicators(design, by = "Sector", approach = params$approach)

by_sector_display <- by_sector[, .(
  Sector,
  `LFPR (%)` = sprintf("%.1f (%.1f-%.1f)", lfpr, lfpr_low, lfpr_upp),
  `WPR (%)` = sprintf("%.1f (%.1f-%.1f)", wpr, wpr_low, wpr_upp),
  `UR (%)` = sprintf("%.1f (%.1f-%.1f)", ur, ur_low, ur_upp),
  N = format(n, big.mark = ",")
)]

kable(by_sector_display,
      caption = "Labour Force Indicators by Sector (95% CI)",
      align = c("l", "r", "r", "r", "r"))
```

## By Sex and Sector

```{r national-by-sex-sector}
#| label: national-by-sex-sector

by_sex_sector <- calc_all_indicators(design, 
                                      by = c("Sex", "Sector"), 
                                      approach = params$approach)

by_sex_sector_display <- by_sex_sector[, .(
  Sex, Sector,
  `LFPR (%)` = round(lfpr, 1),
  `WPR (%)` = round(wpr, 1),
  `UR (%)` = round(ur, 1)
)]

kable(by_sex_sector_display,
      caption = "Labour Force Indicators by Sex and Sector",
      align = c("l", "l", "r", "r", "r"))
```

```{r plot-sex-sector}
#| label: plot-sex-sector
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "LFPR by Sex and Sector"

ggplot(by_sex_sector, aes(x = Sector, y = lfpr, fill = Sex)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = lfpr_low, ymax = lfpr_upp),
                position = position_dodge(width = 0.8),
                width = 0.2) +
  geom_text(aes(label = sprintf("%.1f", lfpr)),
            position = position_dodge(width = 0.8),
            vjust = -1.5, size = 3) +
  scale_fill_pub("binary") +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Labour Force Participation Rate by Sex and Sector",
    subtitle = sprintf("PLFS %s, %s approach, Age 15+", 
                       params$survey_year,
                       if(params$approach == "ps") "Principal Status" else "CWS"),
    x = NULL,
    y = "LFPR (%)",
    fill = NULL,
    caption = "Error bars show 95% confidence intervals"
  ) +
  theme_publication()
```

---

# State-wise Analysis

```{r state-indicators}
#| label: state-indicators

by_state <- calc_indicators_by_state(design, approach = params$approach, add_names = TRUE)

# Sort by LFPR
setorder(by_state, -lfpr)

# Display top 10 and bottom 10
cat("### Top 10 States by LFPR\n")
kable(by_state[1:10, .(state_name, 
                        `LFPR (%)` = round(lfpr, 1),
                        `WPR (%)` = round(wpr, 1),
                        `UR (%)` = round(ur, 1),
                        N = format(n, big.mark = ","))],
      caption = "Top 10 States by LFPR")
```

```{r state-bottom}
#| label: state-bottom

cat("### Bottom 10 States by LFPR\n")
n_states <- nrow(by_state)
kable(by_state[(n_states-9):n_states, .(state_name, 
                                         `LFPR (%)` = round(lfpr, 1),
                                         `WPR (%)` = round(wpr, 1),
                                         `UR (%)` = round(ur, 1),
                                         N = format(n, big.mark = ","))],
      caption = "Bottom 10 States by LFPR")
```

```{r plot-state-lfpr}
#| label: plot-state-lfpr
#| fig-width: 8
#| fig-height: 10
#| fig-cap: "LFPR by State"

# All states horizontal bar
p_state <- plot_horizontal_bars(
  by_state[!is.na(state_name)],
  x_var = "lfpr",
  y_var = "state_name",
  title = sprintf("Labour Force Participation Rate by State, PLFS %s", params$survey_year),
  x_label = "LFPR (%)",
  highlight_top = 5
)
p_state
```

---

# Employment Distribution

```{r employment-type}
#| label: employment-type

emp_dist <- calc_employment_distribution(design, approach = params$approach)

emp_dist_display <- emp_dist[!is.na(employment_type), .(
  `Employment Type` = employment_type,
  `Share (%)` = round(share, 1),
  N = format(n, big.mark = ",")
)]

kable(emp_dist_display,
      caption = "Distribution of Workers by Employment Type",
      align = c("l", "r", "r"))
```

```{r employment-by-sex}
#| label: employment-by-sex

emp_by_sex <- calc_employment_distribution(design, by = "Sex", approach = params$approach)

emp_by_sex_wide <- dcast(emp_by_sex[!is.na(employment_type)], 
                          employment_type ~ Sex, 
                          value.var = "share")

kable(emp_by_sex_wide,
      caption = "Employment Type Distribution by Sex (%)",
      digits = 1,
      align = c("l", "r", "r"))
```

```{r plot-employment}
#| label: plot-employment
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Employment Type Distribution by Sex"

ggplot(emp_by_sex[!is.na(employment_type)], 
       aes(x = employment_type, y = share, fill = Sex)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", share)),
            position = position_dodge(width = 0.8),
            vjust = -0.5, size = 3) +
  scale_fill_pub("binary") +
  scale_y_continuous(limits = c(0, 70), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Employment Type Distribution by Sex",
    subtitle = sprintf("PLFS %s, Among employed persons (Age 15+)", params$survey_year),
    x = NULL,
    y = "Share (%)",
    fill = NULL
  ) +
  theme_publication() +
  theme(axis.text.x = element_text(angle = 0))
```

---
  
# Export Tables

```{r export-tables}
#| label: export-tables
#| eval: false

# Export all tables to Word and LaTeX
export_indicator_table(by_sex, 
                       sprintf("lfpr_by_sex_%s", gsub("-", "_", params$survey_year)),
                       title = sprintf("Labour Force Indicators by Sex, PLFS %s", params$survey_year))

export_indicator_table(by_state,
                       sprintf("lfpr_by_state_%s", gsub("-", "_", params$survey_year)),
                       title = sprintf("Labour Force Indicators by State, PLFS %s", params$survey_year))

message("Tables exported to outputs/tables/")
```

---

# Technical Notes

## Data Source

- **Survey**: Periodic Labour Force Survey (PLFS) `r params$survey_year`
- **Conducted by**: National Statistical Office (NSO), MoSPI, Government of India
- **Sample Design**: Stratified multi-stage sampling
- **Reference Period**: July `r substr(params$survey_year, 1, 4)` to June `r substr(params$survey_year, 6, 7)`

## Definitions

- **LFPR**: Persons in labour force (employed + unemployed) as percentage of population
- **WPR**: Employed persons as percentage of population  
- **UR**: Unemployed persons as percentage of labour force
- **Principal Status (PS)**: Activity status based on major time spent during 365 days preceding survey
- **Current Weekly Status (CWS)**: Activity status based on 7 days preceding survey

## Activity Status Codes

- **Employed (11-51)**: Self-employed, regular wage/salaried, casual labour
- **Unemployed (61)**: Seeking or available for work
- **Not in Labour Force (71-98)**: Students, domestic duties, others

---

*Report generated on `r Sys.Date()` using the IndiaData analysis system.*
